Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular cross sections

### 2011-09-23 Text/Submission/paper.tex
Almost finished a referee report.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works on three-dimensional flows. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition of flow regime in a liquid
slug. Vortexes observed at lower capillary numbers disappear as the capillary number goes beyond a
certain threshold,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to collapse results of radii dependency on the capillary number
simulations of rectangular channels with different aspect ratios on a single curve.
It was also found that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
circular for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the
relative slug to bubble velocity for
a range of capillary numbers.

@@

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. The method is a particle method which allows one to simulate physical
phenomena on the microscopic level. For instance, the introduction of the force or potential on the
microscopic level allows to restore multiphase macroscopic equations \cite{swift,
Shan-chen:extended}.

The binary liquid free-energy LB model due to \citet{swift} that we used
simulates two liquids with the assumption of uniform overall
density. %As it was indicated by \citet{gupta-review} a flow in microchannels is a complex
%interplay of viscous, gravitational, inertial and surface forces. 
The classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different density and viscosity. {\color{red} The system is simulated in a right manner if all
non-dimensional parameters are matched in real physical system and simulation system.
However, one needs to consider the leading non-dimensional numbers which are responsible for
physical phenomenon. Even, the binary-liquid model cannot match the gas-liquid density ratio, we
consider it as a non-leading parameter. First of all, the numerical
results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers, $Re<70$. The parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. The maximum attained
Reynolds number $Re<10$. The second reason is that density ratio is involved while describing a
bubble shape change. There is a corresponding number defining the shape change, Bond number. Bond
number is $Bo=\frac{Ca\,Re}{Fr^2}$. For real microchannel systems the Bond number is less than
$Bo<0.1$ and we can neglect a bubble shape change.}
Therefore, a major governing
factor for microchannel flows is not the density ratio, but
the
viscosity ratio. As a result, the uniform density binary liquid model is suitable for this kind of
simulations.  
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

@@

\section{Numerical benchmark approach}
\label{sec:numerical:benchmark}
The main discussion here is based on the two-dimensional benchmark approach
\cite{kuzmin-binary2d}. It was indicated that the benchmark layout should have certain
dimensions to conduct simulations. The classical Bretherton benchmark layout is represented in
Fig.~\ref{fig:classical:benchmark}. It describes the gas finger propagation through the liquid
medium.
The film thickness in this case is measured at the inlet. In the lattice Boltzmann framework, such a
formulation has certain challenges \cite{kuzmin-binary2d}. Some of them are attributed to the
dynamic coupling of the inlet and outlet conditions \cite{giavedoni-numerical}. Instead, we
proposed the numerical benchmark indicated in Fig.~\ref{fig:lbm:benchmark}. The dimensions of the
channel are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}} \times 15 H_{\mathrm{eff}}$. The
initial bubble length is taken as $5 H_{\mathrm{eff}}$. \citet{giavedoni-numerical} showed that
the film in the two-dimensional geometry stabilizes at distances of $2.6-4.0$ diameters
from the front tip depending on the Reynolds number. \citet{heil-threedim} measured the bubble radii
at the distance of $5.5 H_{\mathrm{eff}}$ and indicated it to be sufficient for $Ca\leq 10$. In
this work, the following relation holds for all conducted simulations: $0.05 \geq Ca \leq 6$.
Thus, the
film thickness is chosen to be measured in the
middle of the bubble, which is located at least at a distance of $2.5-3$ channel heights from the 
bubble tip.%at the distances one channel height from the rear tip upto one channel

@@

\citet{heil-threedim} and Fig. \ref{fig:heil:three:dim} specify that the diagonal bubble radius is
different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.03$. The radii are approximately the same for
larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations.

@@

The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken to be linearly dependent on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}(\tau_{\mathrm{liq}}-\tau_{\mathrm{gas}})$. This
makes it possible to smoothly change viscosity from the gas viscosity
$\nu_{\mathrm{gas}}=\frac{1}{3}\Bigl(\tau_{\mathrm{gas}}-\frac{1}{2}\Bigr)$ to the liquid viscosity
$\nu_{\mathrm{liq}}=\frac{1}{3}\Bigl(\tau_{\mathrm{liq}}-\frac{1}{2}\Bigr)$ in a region where a
phase gradient exists. The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k
A}{9}}$.

\section{Results}
\label{sec:results}
This section describes numerical simulations. We refer to \ref{append:scaling} for
initialization and scaling procedures. The simulations were conducted on the following 
fluid domain grids $100 \times 100 \times 1500$, $160 \times 160 \times 1500$,
$160 \times 160 \times 2400$ and $200 \times 200 \times 3000$. Body forces were varied as
$\frac{\mathrm{d}P}{\mathrm{d}z}=10^{-6}-10^{-4}$ lattice units. The binary-liquid model parameters
were kept in
the stable region $k=A=0.004,0.04$. The relaxation rates were taken as $\tau_{\mathrm{liq}}=2.5$
and $\tau_{\mathrm{gas}}=0.7$ giving a gas-over-liquid viscosity ratio of $10$. The relaxation
parameter for the phase field was $\tau_{\phi}=1.0$.

@@

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

@@

\subsection{Variation over bubble length}
The work of \citet{wang-non-circular} shows the variation of the bubble
radii along the
bubble. The bubble shape in terms of axial and diagonal radii is reconstructed for a number of
capillary numbers, Fig. \ref{fig:bubble:variation:capillaries}. One can see that for smaller
capillary numbers the diagonal radius exhibits a small jump near the rear bubble cap. That agrees
with results of \citet{wang-non-circular}. However, our simulations show smoother behavior of
the jump.  
\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{bubble_rad_axis.eps}\\
\includegraphics[width=0.8\textwidth]{bubble_rad_diag.eps}\\
\caption{Radii variations along the bubble for different capillary numbers in the plane $x=0$.
One can see that the diagonal radius shows the jump near the bubble rear cap. This agrees with the
VOF simulations of \citet{wang-non-circular}.Coordinate $z$ increases in the streamwise direction.
\label{fig:bubble:variation:capillaries}}
\end{figure}

Another interesting phenomenon regarding the bubble shape was indicated by \citet{heil-threedim}. In
their simulations the non-axisymmetric shape was observed even for larger capillary numbers $Ca>4$
near
the bubble front cap which after certain distance relaxes to have radii equal to each other. The
bubble shape was calculated for the large capillary number $Ca=6.43$, see Fig.
\ref{fig:bubble:ca:large}. However, even the diagonal radius shows the jump near the front tip of
the bubble and it resembles the shape given by \citet{heil-threedim}, but the difference is around
$1$ percent and can be explained by error in the linear interpolation used for the
interface
calculation. Therefore, we do not observe the symmetry breakage for $Ca>4$. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{bubble_ca_large.eps}
\caption{The bubble shape for the capillary number $Ca=6.43$ is presented. $z$ increases in the
streamwise direction. One can see that near the front tip (large $z$) the diagonal radius exhibits
a jump which is relaxed after to the axis radius. Even this resembles
the shape of the curves obtained by \citet{heil-threedim}, but the
difference between radii is around $1$ percent and can be attributed to
the linear interpolation used in the calculation of the interface.\label{fig:bubble:ca:large}}
\end{figure}

\subsection{Capillary number}
The purpose of this section is to study dependence of the radii on the capillary number. The film
thicknesses were extracted from the middle of the bubble. The data presented in Fig.
\ref{fig:capillary:comparison} is simulated with different techniques including different
initialization, and different grids: $52 \times 52 \times 1500$, $82 \times 82 \times 1500$ and $82
\times 82 \time 2400$. All our results are consistent. Therefore one can conclude that for the
capillary number $Ca\geq
0.05$ grid independence is achieved. The results are compared with the results of
\citet{heil-threedim} and of \citet{wang-non-circular}. The data were extracted from the
corresponding references with the help of the program ``Engauge Digitizer''. We observe that
lattice Boltzmann simulations exhibit 
the transition between non-axisymmetric and axisymmetric shape of the bubble.

@@

\subsection{Relative velocity}
One of the differencies between the flow in tubes and square capillaries is that the flow has
different streamline patterns. In square microchannels, the main mass flow occurs through corners
\cite{heil-threedim,wang-non-circular}. This affects the mass flow significantly. One of the
characteristics of the mass flow is the normalized dependence of the relative velocity on the
capillary
number. The bubble always moves faster than the surrounding liquid. The following quantity is
defined as
relative velocity \cite{cerro-bubble-train}:
\begin{equation}
W=\frac{U_{\mathrm{bubble}}-U_{ls}}{U_{\mathrm{bubble}}},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble interface velocity, and $U_{ls}$ is the liquid superficial
velocity. As far as the steady state is achieved one can take the liquid superficial velocity
exactly at the cross section in the middle between bubbles. The liquid superficial velocity is
defined as:
\begin{equation}
U_{ls}=\frac{\int_{A}{U_{liq}dA}}{A}.
\end{equation}
The relative velocity was calculated and compared with the results of \citet{wang-non-circular},
see Fig. \ref{fig:relative:velocity}. One can see an agreement between the simulations indicating
that
one of the mass transport characteristics can be captured accurately. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{relative_velocity.eps}
\caption{Relative velocity comparison between current simulations and simulations of
\citet{wang-non-circular}. One can see a qualitative agreement. \label{fig:relative:velocity}}
\end{figure}

>>>>

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. $r_h$ and $r_d$ are scaled on
the half side length $H_{\mathrm{eff}}/2$. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works on three-dimensional flows. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition of flow regime in a liquid
slug. Vortexes observed at lower capillary numbers disappear as the capillary number goes beyond a
certain threshold,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to collapse results of radii dependency on the capillary number
for rectangular channels with different aspect ratios on a single curve.
It was also found that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
circular for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the
relative slug to bubble velocity for
a range of capillary numbers.

@@

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. The method is a particle method which allows one to simulate physical
phenomena on the microscopic level. For instance, the introduction of the force or potential on the
microscopic level allows to restore multiphase macroscopic equations \cite{swift,
Shan-chen:extended}. {\color{red} Nowadays the method is finding its way to more practical
applications, for example magnetically driven droplet break-up and vaporization
\cite{falcucci-vapour}, mixing in cross-channel flow \cite{jos-mixing}, etc. }

The binary liquid free-energy LB model due to \citet{swift} that we used
simulates two liquids with the assumption of uniform overall
density. 
 %As it was indicated by \citet{gupta-review} a flow in microchannels is a complex
%interplay of viscous, gravitational, inertial and surface forces. 
The classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different density and viscosity. {\color{red} The system is simulated in a right manner if all
non-dimensional parameters are matched in real physical system and simulation system.
However, one needs to consider the leading non-dimensional numbers which are responsible for
physical phenomenon. Even, the binary-liquid model cannot match the gas-liquid density ratio, we
consider it as a non-leading parameter. First of all, the numerical
results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers, $Re<70$. The parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. The maximum attained
Reynolds number $Re<10$. The second reason is that density ratio is involved while describing a
bubble shape change. There is a corresponding number defining the shape change, Bond number. Bond
number is $Bo=\frac{Ca\,Re}{Fr^2}$. For real microchannel systems the Bond number is less than
$Bo<0.1$ and we can neglect a bubble shape change.}
Therefore, a major governing
factor for microchannel flows is not the density ratio, but
the
viscosity ratio. As a result, the uniform density binary liquid model is suitable for this kind of
simulations.

{\color{red} Note that other multiphase models can be used which can account for different
densities, as the pseudopotential based approach \cite{Sbragaglia-Succi} or color model
\cite{swift}. In the case of the pseudopotential based approach, one needs to use the
binary-liquid formulation as to avoid possible condensation/evaporation \cite{pooley-mrt-filling}
and to perform Laplace law simulations to obtain the surface tension as it is not given explicitly.
For the case of the color model, one needs to use more complicated schemes to reconstruct the
surface. All other finding of the paper can be applicable for all the abovementioned models.}
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

@@

\section{Numerical benchmark approach}
\label{sec:numerical:benchmark}
The main discussion here is based on the two-dimensional benchmark approach
\cite{kuzmin-binary2d}. It was indicated that the benchmark layout should have certain
dimensions to conduct simulations. The classical Bretherton benchmark layout is represented in
Fig.~\ref{fig:classical:benchmark}. It describes the gas finger propagation through the liquid
medium.
The film thickness in this case is measured at the inlet. In the lattice Boltzmann framework, such a
formulation has certain challenges \cite{kuzmin-binary2d}. Some of them are attributed to the
dynamic coupling of the inlet and outlet conditions \cite{giavedoni-numerical}. Instead, we
proposed the numerical benchmark indicated in Fig.~\ref{fig:lbm:benchmark}. The dimensions of the
channel are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}} \times 15 H_{\mathrm{eff}}$. The
initial bubble length is taken as $5 H_{\mathrm{eff}}$. \citet{giavedoni-numerical} showed that
the film in the two-dimensional geometry stabilizes at distances of $2.6-4.0$ diameters
from the front tip depending on the Reynolds number. \citet{heil-threedim} measured the bubble radii
at the distance of $5.5 H_{\mathrm{eff}}$ and indicated it to be sufficient for $Ca\leq 10$. In
this work, the following relation holds for all conducted simulations: $0.05 \leq Ca \leq 6$.
Thus, the
film thickness is chosen to be measured in the
middle of the bubble, which is located at least at a distance of $2.5-3$ channel heights from the 
bubble tip.%at the distances one channel height from the rear tip upto one channel

@@

\citet{heil-threedim} and Fig. \ref{fig:heil:three:dim} specify that the diagonal bubble radius is
different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.03$. The radii are approximately the same for
larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations. {\color{red} However, as it will
be shown later the film thickness in simulations is larger than the example number ($0.49
H_{\mathrm{eff}}$ for $Ca\approx 0.03$). Therefore, actual grids are smaller (indicated in
Section \ref{sec:results}) than the computationally demanding grid $600 \times 600 \times 9000$.}

@@

The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken to be linearly dependent on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}(\tau_{\mathrm{liq}}-\tau_{\mathrm{gas}})$. This
makes it possible to smoothly change viscosity from the gas viscosity
$\nu_{\mathrm{gas}}=\frac{1}{3}\Bigl(\tau_{\mathrm{gas}}-\frac{1}{2}\Bigr)$ to the liquid viscosity
$\nu_{\mathrm{liq}}=\frac{1}{3}\Bigl(\tau_{\mathrm{liq}}-\frac{1}{2}\Bigr)$ in a region where a
phase gradient exists. The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k
A}{9}}$. {\color{red} The walls were simulated using the bounce-back rule \cite{yu}. The walls were
taken with neutral wettability as results do not depend on the wettability \cite{kuzmin-binary2d}.}

\section{Results}
\label{sec:results}
This section describes numerical simulations. We refer to \ref{append:scaling} for
initialization and scaling procedures. The simulations were conducted on the following 
fluid domain grids $100 \times 100 \times 1500$, $160 \times 160 \times 1500$,
$160 \times 160 \times 2400$ and $200 \times 200 \times 3000$. Body forces were varied as
$\frac{\mathrm{d}P}{\mathrm{d}z}=10^{-6}-10^{-4}$ lattice units.{\color{red} The binary-liquid model
parameters
were kept in
the stable region $k=A=0.04$ for most simulations and $k=A=0.004$ for simulations with $Ca=6.70$. 
Note that the choice of $k=A$ guarantee that the film thickness is $5$ grid spacings, which is a
good choice for stability issues. } The relaxation rates were taken as $\tau_{\mathrm{liq}}=2.5$
and $\tau_{\mathrm{gas}}=0.7$ giving a gas-over-liquid viscosity ratio of $10$. The relaxation
parameter for the phase field was $\tau_{\phi}=1.0$.

@@

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[htb!]
\includegraphics[width=0.8\textwidth]{phase_crossection_ca5.eps}\\
\includegraphics[width=0.8\textwidth]{phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(top) and $Ca=1.13$ (bottom). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

@@

\subsection{Variation over bubble length}
The work of \citet{wang-non-circular} shows the variation of the bubble
radii along the
bubble. The bubble shape in terms of axial and diagonal radii is reconstructed for a number of
capillary numbers, Fig. \ref{fig:bubble:variation:capillaries}. One can see that for smaller
capillary numbers the diagonal radius exhibits a small jump near the rear bubble cap. That agrees
with results of \citet{wang-non-circular}. However, our simulations show smoother behavior of
the jump.  
\begin{figure}[htb!]
\includegraphics[width=0.45\textwidth]{bubble_rad_axis.eps}\hfill
\includegraphics[width=0.45\textwidth]{bubble_wang_axis.eps}\\
\includegraphics[width=0.45\textwidth]{bubble_rad_diag.eps}\hfill
\includegraphics[width=0.45\textwidth]{bubble_wang_diag.eps}\\
\caption{Simulations radii variations (left)  along the bubble for different capillary numbers in
the plane $x=0$.
One can see that the diagonal radius shows the jump near the bubble rear cap. This agrees with the
VOF simulations of \citet{wang-non-circular} (right).Coordinate $z$ increases in the streamwise
direction. The discretization of \citet{wang-non-circular} data was performed with ''Engauge
Digitizer``.
\label{fig:bubble:variation:capillaries}}
\end{figure}

Another interesting phenomenon regarding the bubble shape was indicated by \citet{heil-threedim}. In
their simulations the non-axisymmetric shape was observed even for larger capillary numbers $Ca>4$
near
the bubble front cap which after certain distance relaxes to have radii equal to each other. The
bubble shape was calculated for the large capillary number $Ca=6.43$, see Fig.
\ref{fig:bubble:ca:large}. However, even the diagonal radius shows the jump near the front tip of
the bubble and it resembles the shape given by \citet{heil-threedim}, but the difference is around
$1$ percent and can be explained by error in the linear interpolation used for the
interface
calculation. Therefore, we do not observe the symmetry breakage for $Ca>4$. 
\begin{figure}[ht]
\includegraphics[width=0.9\textwidth]{bubble_ca_large.eps}\\
\includegraphics[width=0.9\textwidth]{bubble_heil_ca10.eps}
\caption{The bubble shape for the capillary number $Ca=6.43$ is presented (top). $z$ increases in
the
streamwise direction. One can see that near the front tip (large $z$) the diagonal radius exhibits
a jump which is relaxed to the axis radius (small $z$).{\color{red} Even this resembles
the shape of the curves obtained by \citet{heil-threedim} (bottom, opposite streamwise direction),
but the
difference between radii is around $1$ percent and can be attributed to
the linear interpolation used in the calculation of the interface.}\label{fig:bubble:ca:large}}
\end{figure}

\subsection{Capillary number}
The purpose of this section is to study dependence of the radii on the capillary number. The film
thicknesses were extracted from the middle of the bubble. The data presented in Fig.
\ref{fig:capillary:comparison} is simulated with different techniques including different
initialization, and different grids: $52 \times 52 \times 1500$, $82 \times 82 \times 1500$ and $82
\times 82 \times 2400$. All our results are consistent {\color{red} and show the grid and
initialization independency}. Therefore one can conclude that for the
capillary number $Ca\geq
0.05$ grid independence is achieved. The results are compared with the results of
\citet{heil-threedim} and of \citet{wang-non-circular}. The data were extracted from the
corresponding references with the help of the program ``Engauge Digitizer''. We observe that
lattice Boltzmann simulations exhibit 
the transition between non-axisymmetric and axisymmetric shape of the bubble.

@@

\subsection{Relative velocity}
One of the differences between the flow in tubes and square capillaries is that the flow has
different streamline patterns. In square microchannels, the main mass flow occurs through corners
\cite{heil-threedim,wang-non-circular}. This affects the mass flow significantly. One of the
characteristics of the mass flow is the normalized dependence of the relative velocity on the
capillary
number. The bubble always moves faster than the surrounding liquid. The following quantity is
defined as
relative velocity \cite{cerro-bubble-train}:
\begin{equation}
W=\frac{U_{\mathrm{bubble}}-U_{ls}}{U_{\mathrm{bubble}}},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble interface velocity, and $U_{ls}$ is the liquid superficial
velocity. As far as the steady state is achieved one can take the liquid superficial velocity
exactly at the cross section in the middle between bubbles. The liquid superficial velocity is
defined as:
\begin{equation}
U_{ls}=\frac{\int_{A}{U_{liq}dA}}{A}.
\end{equation}
The relative velocity was calculated and compared with the results of \citet{wang-non-circular},
see Fig. \ref{fig:relative:velocity}. One can see an agreement between the simulations indicating
that
one of the mass transport characteristics can be captured accurately. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{relative_velocity.eps}
\caption{Relative velocity comparison between current simulations and simulations of
\citet{wang-non-circular}. One can see a qualitative agreement. \label{fig:relative:velocity}}
\end{figure}
